\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-09 paper.qmd
Selected another application with mouse barin data

### Non-parametric models

#### Hexagonal binning

Hexagonal binning is a data visualization technique that aggregates high-dimensional data into a two-dimensional representation using hexagonal regions called bins [@article66]. The hexagonal shape is preferred due to its circular-like appearance, providing smoother transitions between neighboring bins and efficient data aggregation. This technique is especially useful for handling large data sets as it reduces visual clutter while capturing the underlying data distribution effectively. By considering the data distribution, important features are captured, and gaps between bins are minimized, leading to a more accurate representation of the data. In our algorithm, hexagonal binning is utilized to create a low-dimensional manifold and facilitate data visualization in the original high-dimensional space.

#### Triangulation

Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30] finding applications in mesh generation and representing complex geometric structures efficiently and discretely. One common algorithm used for triangulation is Delaunay triangulation [@article26], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

In our algorithm, we are interested in triangulation as it plays a key role in providing a clear and concise representation of the data. By converting the data points into interconnected triangles, we can visualize and analyze the data in a more structured and organized manner. Triangulation helps us identify geometric relationships and patterns in the data, enabling us to make informed decisions and gain valuable insights from the information presented in the triangular form.

@@

@@

In this step, the algorithm proceeds to triangulate the centroids (see @fig-meth (Step 3)) of the hex bins using the `tripack` package in R [@article54]. Since we are working with the centroids of regular hexagonal bins, the resulting mesh will predominantly comprise equal-sized regular triangles. However, the triangulation also helps span any gaps that may exist between clusters of points, allowing for a more complete and interconnected representation of the data.

@@

a1 + a2 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```
### Tunning the model

### Summaries of the model {#sec-diagnostics}

@@

In our prediction process, we employ the K-nearest neighbors (KNN) algorithm to predict the hex bin IDs. Specifically, for each new high-dimensional data point, the algorithm assigns the first neighboring hex bin ID as the predicted hex bin ID.

<!-- move to goodness of fit stat-->
Moving on to the assessment of prediction accuracy, we calculate the Mean Squared Error (MSE). The MSE helps measure the average squared differences between the actual high-dimensional data ($x_{ij}$) and the predicted bin centroid data in high-D ($C_{x_ij}$) values across all bins. Mathematically, this is expressed as:

@@

Here, $b$ signifies the total number of bins, $p$ denotes the number of dimensions in the high-dimensional (high-D) data, and $n$ represents the number of observations.

@@

In this analysis, we're looking at how well our model is doing in two aspects. First, we're checking the distances in our lower-dimensional space (see @fig-diag (a)). We want to see if the distances between different types of edges are similar. If they are, it means our model is doing a good job of keeping the relationships consistent. If there are big differences, it could mean our model is struggling with certain types of edges.

@@

<!--### Software

@@

### MNIST

The MNIST dataset is a widely recognized benchmark in the realm of computer vision and machine learning, which is a collection of 28x28 pixel grayscale images of handwritten digits (0 through 9), serving as a foundational resource for developing and evaluating various image classification algorithms. With its simplicity and clarity, MNIST provides an ideal platform for researchers and practitioners to experiment with and benchmark their models.

>>>>

@@

Hexagonal binning is a data visualization technique that aggregates high-dimensional data into a two-dimensional representation using hexagonal regions called bins [@article66]. The hexagonal shape is preferred due to its circular-like appearance, providing smoother transitions between neighboring bins and efficient data aggregation. This technique is especially useful for handling large data sets as it reduces visual clutter while capturing the underlying data distribution effectively. By considering the data distribution, important features are captured, and gaps between bins are minimized, leading to a more accurate representation of the data. In our algorithm, hexagonal binning is utilized to create a low-dimensional manifold.

@@

In this step, the algorithm proceeds to triangulate the centroids (see @fig-meth (Step 3)) of the hex bins. Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation. In our algorithm, triangulation helps us identify geometric relationships and patterns in the data.

Since we are working with the centroids of regular hexagonal bins, the resulting mesh will predominantly comprise equal-sized regular triangles. However, the triangulation also helps span any gaps that may exist between clusters of points, allowing for a more complete and interconnected representation of the data.

@@

a1 + a2 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

### Tunning the model

### Summaries of the model {#sec-summary}

@@

In this context, the term "prediction values" refers to the 2D coordinates predicted for the NLDR technique. The approach involves employing the K-nearest neighbors (KNN) algorithm to identify the nearest hexagonal bin centroid in the 2D space. Subsequently, the coordinates of this centroid are assigned as the low-dimensional predicted values for the test data in 2D space. It is noteworthy that traditional NLDR methods, such as t-SNE, often lack a direct predict function, making our approach valuable for generating predicted values in the absence of such functionalities.

The concept of "residuals" is pivotal in evaluating the accuracy of representing bin centroids in high dimensions. To quantify this accuracy, we introduce an error metric, which measures the sum of squared differences between the high-dimensional data (\(x_{ij}\)) and the predicted bin centroid data in high-dimensional space (\(C_{x_ij}\)) across all bins and dimensions. Mathematically, this error is expressed as:

$$
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
$$ {#eq-equation11}

Here, \(n\) represents the number of bins, \(p\) represents the dimensions, \(x_{ij}\) is the actual high-dimensional data, and \(C_{x_ij}\) is the predicted bin centroid data in high dimensions.

The error metric outlined above provides valuable insights into the overall accuracy of our predictive model. By quantifying the squared deviations between the actual and predicted values across all bins and dimensions, we gain a comprehensive understanding of how well our method captures and represents the underlying structure of the data in the reduced 2D space. This assessment is crucial for evaluating the efficacy of our NLDR technique in preserving the essential information present in the original high-dimensional data.

Moving on to the assessment of prediction accuracy, we calculate the Mean Squared Error (MSE). The MSE helps measure the average squared differences between the actual high-dimensional data ($x_{ij}$) and the predicted bin centroid data in high-D ($C_{x_ij}$) values across all bins. Mathematically, this is expressed as:

@@

Here, $b$ signifies the total number of bins, $p$ denotes the number of dimensions in the high-dimensional data, and $n$ represents the number of observations.

@@

<!--In this analysis, we're looking at how well our model is doing in two aspects. First, we're checking the distances in our lower-dimensional space (see @fig-diag (a)). We want to see if the distances between different types of edges are similar. If they are, it means our model is doing a good job of keeping the relationships consistent. If there are big differences, it could mean our model is struggling with certain types of edges.

@@

### Software

@@

### Single-Cell Tagged Reverse Transcription sequencing data of mouse

The Zeisel mouse brain dataset, obtained through Spatial Transcriptomics (STRT-Seq). Within this dataset, information is collected from a substantial 2,816 individual mouse brain cells. Each of these cells acts as a molecular snapshot, capturing the distinctive genetic activity within various cell types. This diversity spans neurons, glial cells, and other essential components of the brain, offering a comprehensive view of the cellular tapestry.

What makes this dataset particularly valuable is its ability to shed light on the spatial distribution of cells. Researchers can explore how gene expression patterns vary across different regions of the mouse brain, unlocking insights into the functional specialization of these regions and the intricate networks that underpin neural processes.

### 2024-01-09 paper.tex
Selected another application with mouse barin data

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{non-parametric-models}{%
\subsection{Non-parametric models}\label{non-parametric-models}}

\hypertarget{hexagonal-binning}{%
\subsubsection{Hexagonal binning}\label{hexagonal-binning}}

Hexagonal binning is a data visualization technique that aggregates
high-dimensional data into a two-dimensional representation using
hexagonal regions called bins \citep{article66}. The hexagonal shape is
preferred due to its circular-like appearance, providing smoother
transitions between neighboring bins and efficient data aggregation.
This technique is especially useful for handling large data sets as it
reduces visual clutter while capturing the underlying data distribution
effectively. By considering the data distribution, important features
are captured, and gaps between bins are minimized, leading to a more
accurate representation of the data. In our algorithm, hexagonal binning
is utilized to create a low-dimensional manifold and facilitate data
visualization in the original high-dimensional space.

\hypertarget{triangulation}{%
\subsubsection{Triangulation}\label{triangulation}}

Triangulation is a fundamental process in computational geometry and
computer graphics that involves dividing a set of points in a given
space into interconnected triangles \citep{article30} finding
applications in mesh generation and representing complex geometric
structures efficiently and discretely. One common algorithm used for
triangulation is Delaunay triangulation \citep{article26}, where points
are connected in a way that maximizes the minimum angles of the
resulting triangles, leading to a more regular and well-conditioned
triangulation.

In our algorithm, we are interested in triangulation as it plays a key
role in providing a clear and concise representation of the data. By
converting the data points into interconnected triangles, we can
visualize and analyze the data in a more structured and organized
manner. Triangulation helps us identify geometric relationships and
patterns in the data, enabling us to make informed decisions and gain
valuable insights from the information presented in the triangular form.

@@

@@

In this step, the algorithm proceeds to triangulate the centroids (see
Figure~\ref{fig-meth} (Step 3)) of the hex bins using the
\texttt{tripack} package in R \citep{article54}. Since we are working
with the centroids of regular hexagonal bins, the resulting mesh will
predominantly comprise equal-sized regular triangles. However, the
triangulation also helps span any gaps that may exist between clusters
of points, allowing for a more complete and interconnected
representation of the data.

@@

\hypertarget{sec-diagnostics}{%
\subsection{Summaries of the model}\label{sec-diagnostics}}

@@

In our prediction process, we employ the K-nearest neighbors (KNN)
algorithm to predict the hex bin IDs. Specifically, for each new
high-dimensional data point, the algorithm assigns the first neighboring
hex bin ID as the predicted hex bin ID.

@@

Here, \(b\) signifies the total number of bins, \(p\) denotes the number
of dimensions in the high-dimensional (high-D) data, and \(n\)
represents the number of observations.

@@

In this analysis, we're looking at how well our model is doing in two
aspects. First, we're checking the distances in our lower-dimensional
space (see Figure~\ref{fig-diag} (a)). We want to see if the distances
between different types of edges are similar. If they are, it means our
model is doing a good job of keeping the relationships consistent. If
there are big differences, it could mean our model is struggling with
certain types of edges.

Next, we're examining how our model handles the relationship between
high-dimensional and low-dimensional distances for each type of edge
(see Figure~\ref{fig-diag} (b)). If our data points nicely line up
around the expected line, it tells us that our model is preserving the
original structure well. If they're scattered, it could mean our model
is having trouble with specific data points. This two-part analysis
helps us really understand how our model fits the data.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diag-1.pdf}

\caption{\label{fig-diag}The two diagnostic plot assess the
effectiveness of the high-D representation algorithm performed for the
\textbf{s\_curve\_noise\_umap} data set\(\colon\) (a) visualizing the
differences between low-D distances and (b) visualizing the relationship
between high-D (\(d^{(p)}\)) and low-D distances (\(d^{(2)}\)).}

@@

\hypertarget{sec-applications}{%
\section{Examples/Applications}\label{sec-applications}}

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

@@

Hexagonal binning is a data visualization technique that aggregates
high-dimensional data into a two-dimensional representation using
hexagonal regions called bins \citep{article66}. The hexagonal shape is
preferred due to its circular-like appearance, providing smoother
transitions between neighboring bins and efficient data aggregation.
This technique is especially useful for handling large data sets as it
reduces visual clutter while capturing the underlying data distribution
effectively. By considering the data distribution, important features
are captured, and gaps between bins are minimized, leading to a more
accurate representation of the data. In our algorithm, hexagonal binning
is utilized to create a low-dimensional manifold.

@@

In this step, the algorithm proceeds to triangulate the centroids (see
Figure~\ref{fig-meth} (Step 3)) of the hex bins. Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep{article26}, where points are connected
in a way that maximizes the minimum angles of the resulting triangles,
leading to a more regular and well-conditioned triangulation. In our
algorithm, triangulation helps us identify geometric relationships and
patterns in the data.

Since we are working with the centroids of regular hexagonal bins, the
resulting mesh will predominantly comprise equal-sized regular
triangles. However, the triangulation also helps span any gaps that may
exist between clusters of points, allowing for a more complete and
interconnected representation of the data.

@@

\hypertarget{sec-summary}{%
\subsection{Summaries of the model}\label{sec-summary}}

@@

In this context, the term ``prediction values'' refers to the 2D
coordinates predicted for the NLDR technique. The approach involves
employing the K-nearest neighbors (KNN) algorithm to identify the
nearest hexagonal bin centroid in the 2D space. Subsequently, the
coordinates of this centroid are assigned as the low-dimensional
predicted values for the test data in 2D space. It is noteworthy that
traditional NLDR methods, such as t-SNE, often lack a direct predict
function, making our approach valuable for generating predicted values
in the absence of such functionalities.

The concept of ``residuals'' is pivotal in evaluating the accuracy of
representing bin centroids in high dimensions. To quantify this
accuracy, we introduce an error metric, which measures the sum of
squared differences between the high-dimensional data ((x\_\{ij\})) and
the predicted bin centroid data in high-dimensional space
((C\_\{x\_ij\})) across all bins and dimensions. Mathematically, this
error is expressed as:

\begin{equation}\protect\hypertarget{eq-equation11}{}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\label{eq-equation11}\end{equation}

Here, (n) represents the number of bins, (p) represents the dimensions,
(x\_\{ij\}) is the actual high-dimensional data, and (C\_\{x\_ij\}) is
the predicted bin centroid data in high dimensions.

The error metric outlined above provides valuable insights into the
overall accuracy of our predictive model. By quantifying the squared
deviations between the actual and predicted values across all bins and
dimensions, we gain a comprehensive understanding of how well our method
captures and represents the underlying structure of the data in the
reduced 2D space. This assessment is crucial for evaluating the efficacy
of our NLDR technique in preserving the essential information present in
the original high-dimensional data.

@@

Here, \(b\) signifies the total number of bins, \(p\) denotes the number
of dimensions in the high-dimensional data, and \(n\) represents the
number of observations.

@@

@@

\hypertarget{sec-applications}{%
\section{Applications}\label{sec-applications}}

@@

\hypertarget{zeisel-mouse-brain-strt-seq}{%
\subsection{Zeisel mouse brain
(STRT-Seq)}\label{zeisel-mouse-brain-strt-seq}}

The Zeisel mouse brain dataset, obtained through Spatial Transcriptomics
(STRT-Seq). Within this dataset, information is collected from a
substantial 2,816 individual mouse brain cells. Each of these cells acts
as a molecular snapshot, capturing the distinctive genetic activity
within various cell types. This diversity spans neurons, glial cells,
and other essential components of the brain, offering a comprehensive
view of the cellular tapestry.

What makes this dataset particularly valuable is its ability to shed
light on the spatial distribution of cells. Researchers can explore how
gene expression patterns vary across different regions of the mouse
brain, unlocking insights into the functional specialization of these
regions and the intricate networks that underpin neural processes.
